---
title: "A modern approach to transition analysis and process mining with Markov models: A tutorial with R" 
author: 
   - name: "Jouni Helske"
   - name: "Satu Helske"
   - name: "Mohammed Saqr"
   - name: "Sonsoles López-Pernas"
   - name: "Keefe Murphy"
format: 
  html:
    df-print: "paged"
    fig-align: "center"
dpi: 900
---

## Examples with R
### Preparation
Libraries
```{r, message = FALSE, warning = FALSE, results = FALSE}
library(qgraph)
library(rio)
library(seqHMM)
library(tidyverse)
library(TraMineR)
```

Download data and create sequence

```{r}
URL <- "https://github.com/sonsoleslp/labook-data/raw/main/"
simulated_data <- import(paste0(URL, "12_longitudinalRoles/simulated_roles.csv"))

roles_seq <- seqdef(
  simulated_data, 
  var = 3:22, 
  alphabet = c("Isolate", "Mediator", "Leader"),
  cnames = 1:20
)

cpal(roles_seq) <- c("#FBCE4B", "#F67067", "#5C2262")

cov_data <- simulated_data |>
  select(ID, GPA) |>
  mutate(GPA = factor(GPA, levels = c("Low", "Middle", "High")))
```

### Markov models 


#### Markov model

```{r}
markov_model <- build_mm(roles_seq)
```


```{r}
print(markov_model)
```
We can also draw a graph of the estimated model using the `plot` method which by default shows the states as pie graphs (for the MM, the pie graphs only consist of one state), transition probabilities as directed arrows, and initial probabilities below each state (see @fig-mm-pie).

```{r, fig.width=8, fig.height=3, fig.align="center", fig.cap="Estimated Markov model as a pie charts with the transition probabilities shown as labelled edges."}
#| label: fig-mm-pie
plot(markov_model, 
  legend.prop = 0.2, ncol.legend = 3, 
  edge.label.color = "black", vertex.label.color = "black")
```

#### Hidden Markov models


```{r}
set.seed(1)
hidden_markov_model <- build_hmm(observations = roles_seq, n_states = 2)
```


```{r}
fit_hmm <- fit_model(hidden_markov_model)
fit_hmm$model
```

HMM with two hidden states (pie charts), with transitions between hidden states shown as labelled edges.
```{r, fig.width=8, fig.height=3, fig.align="center"}
#| label: fig-hmm-pie
plot(fit_hmm$model, 
  ncol.legend = 4, legend.prop = 0.2, 
  edge.label.color = "black", vertex.label.color = "black"
)
```

Build model with user-defined starting values
```{r}
## Set seed for randomisation
set.seed(1)

## Initial state probability vector, must sum to one
init_probs <- c(0.3, 0.4, 0.3)

## a 3x3 transition matrix, each row should sum to one
trans_probs <- rbind(c(0.8, 0.15, 0.05), c(0.2, 0.6, 0.2), c(0.05, 0.15, 0.8))

## Simulate emission probabilities
emission_probs <- simulate_emission_probs(
  n_states = 3, n_symbols = length(alphabet(roles_seq))
)

## Build the HMM
hidden_markov_model_2 <- build_hmm(
  roles_seq, initial_probs = init_probs, transition_probs = trans_probs,
  emission_probs = emission_probs
)
```

Fit model
```{r}
set.seed(1)
fit_hmm_2 <- fit_model(hidden_markov_model_2, 
  control_em = list(restart = list(times = 50))
)
```

EM estimation
```{r}
fit_hmm_2$em_results
```

HMM with three hidden states (pie charts), with transitions between hidden states shown as labelled edges.
```{r, fig.width=8, fig.height=4, fig.align="center"}
#| label: fig-hmm-2
plot(fit_hmm_2$model, 
  legend.prop = 0.15, ncol.legend = 3,
  edge.label.color = "black", vertex.label.color = "black",
  combine.slices = 0, trim = 0.0001
)
```

Observed and hidden state sequences from the HMM with three hidden states.
```{r, fig.width=6, fig.height=6, fig.keep = "last", fig.align="center"}
ssplot(fit_hmm_2$model, 
  # Plot sequence index plot (full sequences)
  type = "I", 
  # Plot observed and hidden state sequences
  plots = "both", 
  # Sort sequences by the scores of multidimensional scaling
  sortv = "mds.hidden",
  # X axis tick labels
  xtlab = 1:20
)
```

Build HMM
```{r}
## Set seed for randomisation
set.seed(1)

## Build and estimate a HMM with four states
hidden_markov_model_3 <- build_hmm(roles_seq, n_states = 4)

fit_hmm_3 <- fit_model(hidden_markov_model_3, 
  control_em = list(restart = list(times = 100))
)

fit_hmm_3$em_results$best_opt_restart
```



```{r}
BIC(fit_hmm$model)
BIC(fit_hmm_2$model)
BIC(fit_hmm_3$model)
```


#### Mixture Markov models


```{r}
## Set seed for randomisation
set.seed(123)
## Define model structure (3 clusters)
mmm <- build_mmm(roles_seq, n_clusters = 3)
```


```{r}
fit_mmm <- fit_model(mmm)
```

MMM with three clusters
```{r, fig.width=6, fig.keep = "last", fig.align="center"}
plot(fit_mmm$model, 
  # Plot all clusters at the same time
  interactive = FALSE, 
  # Set the number of rows and columns for cluster plots (one row, three columns)
  nrow = 1, ncol = 3,
  # Omit legends
  with.legend = FALSE, 
  # Choose another layout for the vertices (see plot.igraph)
  layout = layout_in_circle,
  # Omit pie graphs from vertices
  pie = FALSE,
  # Set state colours
  vertex.label.color = c("black", "black", "white"),
  # Set state label colours
  vertex.color = cpal(roles_seq),
  # Increase the size of the circle
  vertex.size = 80,
  # Plot state labels instead of initial probabilities
  vertex.label = "names", 
  # Choose font colour for state labels
  vertex.label.color = "black", 
  # Set state label in the centre of the circle
  vertex.label.dist = 0,
  # Omit labels for transition probabilities
  edge.label = NA
)
```

State distribution plots by most probable clusters estimated with the mixture Markov model (don't forget to click enter on the console)
```{r, fig.width=6, fig.height=4, fig.keep=c(2,3,4), layout=c(2,2), fig.align="center"}
cl1 <- mssplot(fit_mmm$model, 
  # Plot Y axis
  yaxis = TRUE, 
  # Legend position
  with.legend	= "bottom",
  # Legend columns
  ncol.legend = 3,
  # Label for Y axis
  ylab = "Proportion"
)
```


```{r}
set.seed(98765)
mmm_2 <- build_mmm(
  roles_seq, 
  # Starting values for initial probabilities
  initial_probs = fit_mmm$model$initial_probs,
  # Starting values for transition probabilities
  transition_probs = fit_mmm$model$transition_probs,
  # Data frame for covariates
  data = cov_data, 
  # Formula for covariates (one-sided)
  formula = ~ 0 + GPA
)
```


```{r}
#| cache: true
set.seed(12345)
fit_mmm_2 <- fit_model(
  mmm_2, 
  # EM with randomised restarts
  control_em = list(
    restart = list(
      # 50 restarts
      times = 50, 
      # Store loglik values from all 50 + 1 estimation rounds
      n_optimum = 51
    )
  )
)
```


```{r}
fit_mmm_2$em_results$best_opt_restart
```

State distribution plots by most probable clusters estimated with the mixture Markov model with covariates (don't forget to click enter on the console)
```{r, fig.width=6, fig.height=4, fig.keep=c(2,3,4), layout=c(2,2), fig.align="center"}
cluster_names(fit_mmm_2$model) <- c(
  "Mainly leader", "Isolate/mediator", "Mediator/leader"
)
mssplot(fit_mmm_2$model, with.legend = "bottom", ncol.legend = 3)
```


```{r}
summary_mmm_2 <- summary(fit_mmm_2$model)
summary_mmm_2
```


```{r}
exp(fit_mmm_2$model$coefficients)/rowSums(exp(fit_mmm_2$model$coefficients))
```


```{r}
prior_prob <- summary_mmm_2$prior_cluster_probabilities
posterior_prob <- summary_mmm_2$posterior_cluster_probabilities
```

#### Mixture hidden Markov models

```{r, echo = FALSE}
error_hook <- knitr::knit_hooks$get("error")
knitr::knit_hooks$set(error = function(x, options) {
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  error_hook(x, options)
})
```

```{r, linewidth = 80, error = TRUE}
set.seed(123)
mhmm <- build_mhmm(
  roles_seq, 
  n_states = c(2, 2, 2),
  data = cov_data, 
  formula = ~ 0 + GPA
)
fit_mhmm <- fit_model(mhmm)
```


```{r}
set.seed(1)

## Set initial probabilities
init <- list(c(1, 0), c(1, 0), c(1, 0))

## Define own transition probabilities
trans <- matrix(c(
  0.9, 0.1,
  0.1, 0.9
), nrow = 2, byrow = TRUE)

translist <- list(trans, trans, trans)

## Simulate emission probabilities
emiss <- simulate_emission_probs(
  n_states = c(2, 2, 2), 
  n_symbols = 3, 
  n_clusters = 3
)

emiss <- replicate(3, matrix(1/3, 2, 3), simplify = FALSE)

## Define initial values for coefficients
## Here we start from a case where low GPA correlates with Cluster 1, 
## whereas middle and high GPA has no effect
beta <- cbind(0, c(-2, 0, 0), c(-2, 0, 0))

## Define model structure
mhmm_2 <- build_mhmm(
  roles_seq, 
  initial_probs = init, transition_probs = translist, 
  emission_probs = emiss, data = cov_data, 
  formula = ~ 0 + GPA, beta = beta
)
```

Now that we have built the MHMM, we can estimate its parameters:

```{r}
set.seed(1)
suppressWarnings(fit_mhmm_2 <- fit_model(
  mhmm_2,
  control_em = list(restart = list(times = 100, n_optimum = 101)))
)
```

We can now check how many times the log-likelihood values occurred in the 101 estimations:
```{r}
table(round(fit_mhmm_2$em_results$best_opt_restart, 2))
```

MHMM estimated sequence distribution plot with hidden states (don't forget to click enter on the console)

```{r, fig.width=4.6, fig.height=6, fig.keep = c(2,3,4), layout = c(1,1,1), fig.align="center"}
mssplot(fit_mhmm_2$model, 
        plots = "both", type = "I", sortv = "mds.hidden", 
        with.legend = "bottom.combined", legend.prop = .15)
```

Transitions between states for each trajectory

```{r, fig.width=8,fig.height=4 , fig.keep='last', fig.dpi = 100, fig.align='center'}
#| label: fig-mhmm-pie
plot(fit_mhmm_2$model, 
  vertex.size = 60,
  label.color = "black",
  vertex.label.color = "black",
  edge.color = "lightgray",
  edge.label.color = "black",
  legend.prop = 0.4,
  ncol.legend = 1, 
  ncol = 3,
  interactive = FALSE,
  combine.slices = 0
)
```


```{r}
cluster_names(fit_mhmm_2$model) <- c(
  "Downward transition", "Upward transition", "Alternating"
)
```


```{r}
summary(fit_mhmm_2$model)
```


```{r}
exp(fit_mhmm_2$model$coefficients)/rowSums(exp(fit_mhmm_2$model$coefficients))
```

### Stochastic process mining with Markovian models {#process}


```{r, message=FALSE, warning=FALSE}
seq_data <- import(paste0(URL, "1_moodleLAcourse/LMS_data_wide.xlsx"))
seq_data_all <- seqdef(seq_data, var = 7:54 )
```

Sequence index plot for the learning management system logs
```{r, fig.align='center', fig.dpi = 300, fig.width=5, fig.height=5.5, dev='png'}
seqplot(seq_data_all, 
  type = "I", ncol = 4, sortv = "from.start",
  legend.prop = 0.2, cex.legend = 0.7, border = NA,
  ylab = "Sequence (sorted)", xlab = "Time"
)
```

```{r, message = FALSE}
overalltransitions <- seqtrate(seq_data_all)
overalltransitions
```
Process map for the overall process.

```{r, fig.align='center', fig.dpi = 300, fig.width=6, fig.height=6}
Labelx <- alphabet(seq_data_all) # get the labels to use them as nodes names.
transitionsplot <- qgraph(
  overalltransitions, cut = 0.15, minimum = 0.05, 
  labels = Labelx, edge.labels = TRUE, edge.label.cex = 0.65, 
  color = cpal(seq_data_all), curveAll = TRUE, 
  theme = "colorblind", mar = c(4, 3, 4, 3)
)
```


Process map for the overall process with initial probabilities
```{r, fig.align='center', fig.dpi = 300, fig.width=6, fig.height=6}
overallmodel <- build_mm(seq_data_all)

overallplot <- qgraph(
  overalltransitions, 
  cut = 0.15, 
  minimum = 0.05, 
  labels = Labelx, 
  mar = c(4, 3, 4, 3), 
  edge.labels = TRUE, 
  edge.label.cex = 0.65, 
  color = cpal(seq_data_all), 
  curveAll = TRUE, 
  theme = "colorblind", 
  pie = overallmodel$initial_probs
)
```

 

```{r}
set.seed(1)
trans_probs <- simulate_transition_probs(12, 4, diag_c = 5)
init_probs <- as.numeric(prop.table(table(seq_data_all[,1])[1:12]))
init_probs <- replicate(4, init_probs, simplify = FALSE)

builtseqLMS <- build_mmm(
  seq_data_all,
  transition_probs = trans_probs,
  initial_probs = init_probs
)

fitLMS <- fit_model(
  builtseqLMS, 
  global_step = TRUE,
  control_global = list(
    maxtime = 3600, 
    maxeval = 1e5,
    algorithm = "NLOPT_GD_STOGO_RAND"),
  local_step = TRUE,
  threads = 16
)

fitLMS$global_results$message
fitLMS$logLik
```

Extract transition probabilities of each cluster
```{r }
Clustertp1 <- fitLMS$model$transition_probs$`Cluster 1`
Clustertp2 <- fitLMS$model$transition_probs$`Cluster 2`
Clustertp3 <- fitLMS$model$transition_probs$`Cluster 3`
Clustertp4 <- fitLMS$model$transition_probs$`Cluster 4`
```

Extract initial probabilities of each cluster
```{r}
Clusterinitp1 <- fitLMS$model$initial_probs$`Cluster 1`
Clusterinitp2 <- fitLMS$model$initial_probs$`Cluster 2`
Clusterinitp3 <- fitLMS$model$initial_probs$`Cluster 3`
Clusterinitp4 <- fitLMS$model$initial_probs$`Cluster 4`
```

Process maps for each cluster
```{r,fig.width=6, fig.height=5,  fig.align='center', fig.dpi = 300 }
Labelx <- colnames(Clustertp1) # we need to get the labels

Averagelayout <- averageLayout(
  list(Clustertp1, Clustertp2, Clustertp3, Clustertp4)
)
## You can also try with this layout from the previous plot
Overalllayout <- overallplot$layout 

qgraph(
  Clustertp1, cut = 0.15, minimum = 0.05 , labels = Labelx,
  edge.labels = TRUE, edge.label.cex = 0.65, color = cpal(seq_data_all), 
  layout = Averagelayout, pie = Clusterinitp1, curveAll = TRUE, 
  theme = "colorblind", title = "Diverse"
)

qgraph(
  Clustertp2, cut = 0.15, minimum = 0.05, labels = Labelx,
  edge.labels = TRUE, edge.label.cex = 0.65, color = cpal(seq_data_all),  
  layout = Averagelayout, pie = Clusterinitp2, curveAll = TRUE, 
  theme = "colorblind", title = "Assignment-oriented"
)

qgraph(
  Clustertp3, cut = 0.15, minimum = 0.05, labels = Labelx,
  edge.labels = TRUE, edge.label.cex = 0.65, color = cpal(seq_data_all),  
  layout = Averagelayout, pie = Clusterinitp3, curveAll = TRUE, 
  theme = "colorblind", title = "Practical-oriented"
)

qgraph(
  Clustertp4, cut = 0.15, minimum = 0.05 , labels = Labelx, 
  edge.labels = TRUE, edge.label.cex = 0.65, color = cpal(seq_data_all),  
  layout = Averagelayout, pie = Clusterinitp4, curveAll = TRUE, 
  theme = "colorblind", title = "Group-centered"
)
```


```{r}
seq_high <- seq_data_all[seq_data$Achievementlevel4 <= 2,]
seq_low <-  seq_data_all[seq_data$Achievementlevel4 > 2,]

high_mm <- build_mm(seq_high)
low_mm <- build_mm(seq_low)
```

Process maps for high achievers and low achievers using average layout
```{r,fig.width=6, fig.height=2.5, fig.align='center', fig.dpi = 300}
##extract transition probabilities of each cluster
Highprobs <- high_mm$transition_probs
Lowprobs <- low_mm$transition_probs

##extract initial probabilities of each cluster
Highinit <- high_mm$initial_probs
Lowinit <- high_mm$initial_probs

Averagelayout <- averageLayout(list(Highprobs, Lowprobs))

Highplot <- qgraph(
  Highprobs, cut = 0.15, minimum = 0.05, labels = Labelx,
  edge.labels = TRUE, edge.label.cex = 0.65, 
  color = cpal(seq_data_all), layout = Averagelayout, 
  pie = Highinit, theme = "colorblind", title = "High achievers"
)

Lowplot <-  qgraph(
  Lowprobs, cut=0.15, minimum = 0.05, labels = Labelx,
  edge.labels = TRUE, edge.label.cex = 0.65, 
  color = cpal(seq_data_all), layout = Averagelayout, 
  pie = Lowinit, theme = "colorblind", title = "Low achievers"
)
```

Difference between process maps of high achievers and low achievers using average layout
```{r, fig.align='center', fig.dpi = 300}
diffplot <- qgraph(
  Lowprobs - Highprobs, cut = 0.15, minimum = 0.05, labels = Labelx,
  edge.labels = TRUE, edge.label.cex = 0.65, layout = Averagelayout, 
  color = cpal(seq_data_all), theme = "colorblind"
)
```
