---
title: "Structural Equation Modeling with R for Education Scientists"
author:
  - name: "Joran Jongerling"
  - name: "Sonsoles López-Pernas"
  - name: "Mohammed Saqr"
  - name: "Leonie V.D.E. Vogelsmeier"
format: 
  html:
    df-print: "paged"
    fig-align: "center"
---


## SEM in R 

### Preparation

```{r, message = F, warning = F}
library(lavaan) # install.packages("lavaan")
library(tidyverse) # install.packages("tidyverse")
library(rio) # install.packages("rio") 
library(psych) # install.packages("psych")
library(combinat) # install.packages("combinat")
library(devtools) # install.packages("devtools")
library(sleasy) # devtools::install_github("JoranTiU/sleasy")
```

#### Reading in the data

```{r, warning=F}
dataset <- import("https://github.com/lamethods/data/raw/main/4_teachersBurnout/2.%20Response.xlsx")
var_names <- colnames(dataset)
```

Simulating gender covariate
```{r}
set.seed(1611)
dataset$gender <- as.factor((sample(c(rep(0, 618), rep(1, 258)))))
```


#### Are the data suited for SEM?

### Step 1: Steps from the previous chapter

Holdout dataset
```{r}
set.seed(19)
ind <- sample(c(rep("model.building", 438), rep("holdout", 438)))
tmp <- split(dataset, ind)
model.building <- tmp$model.building
holdout <- tmp$holdout
```


### Step 2: Assessing Higher Levels of Invariance

Specify model
```{r}
SEM_model <- '
# Regressing items on factors 
TSC =~ TSC1 + TSC2 + TSC3 + TSC4 + TSC5
TE =~ TE1 + TE2 + TE3 + TE5
EE =~ EE1 + EE2 + EE3 + EE4
DE =~ DE1 + DE2 + DE3
RPA =~ RPA1 + RPA2 + RPA3 + RPA4

# Relations between factors
TE ~ TSC
EE ~ TE + TSC
DE ~ TE + TSC
RPA ~ TE + TSC
'
```

Invariance check
```{r}
invarianceCheck(model = SEM_model, data = holdout, 
                group = "gender", estimator = "MLR", intercept = FALSE, 
                missing = FALSE, display = FALSE)
```


### Step 3: Building the Structural Equation Model and Assessing Fit

Perform the SEM analysis on the holdout data using the specified SEM model 

```{r}
sem_robust <- sem(model = SEM_model, data = holdout, std.lv = TRUE,
              estimator = "MLR", meanstructure = TRUE)
```
 

SEM output
```{r}
sem_structural_results(sem_robust, nd = 3)
```
 
