---
title: "Modeling the Dynamics of Longitudinal Processes in Education. A tutorial with R for The VaSSTra Method"
author:
  - name: "Sonsoles López-Pernas"
  - name: "Mohammed Saqr"
dpi: 900
format: 
  html:
    df-print: "paged"
    fig-align: "center"
---

## *VassTra* with R


### The packages

```{r, results=F, warning=F, message=F}
library(rio)
library(tidyverse)
library(ggplot2)
library(cluster)
library(BBmisc)
library(tidyLPA)
library(TraMineR)
library(seqhandbook)
library(Gmisc)
library(WeightedCluster)
```

### The dataset


```{r}
LongitudinalData <- 
  import("https://github.com/lamethods/data/raw/main/9_longitudinalEngagement/LongitudinalEngagement.csv")
LongitudinalData
```


### From variables to states

Standardize variables
```{r}
LongitudinalData |> group_by(CourseID) |>  
  mutate_at(vars(Freq_Course_View:Active_Days), 
            function(x) normalize(x, method = "standardize")) |> 
  ungroup() -> df
df
```


Select clustering variables
```{r}
to_cluster <- dplyr::select(df, Freq_Course_View, Freq_Forum_Consume, 
                                Freq_Forum_Contribute, Freq_Lecture_View, 
                                Regularity_Course_View, Session_Count, 
                                Total_Duration, Active_Days)
```

Cluster using LPA
```{r, eval=F}
set.seed(22294)
Mclustt <- to_cluster |> ungroup() |> 
  single_imputation() |> 
  estimate_profiles(1:10, models = c(1, 2, 3, 6)) 
```


Compare statistics
```{r, eval=F}
cluster_statistics <- Mclustt |>  
  compare_solutions(statistics = c("AIC", "BIC","Entropy"))
fits <- cluster_statistics$fits |> mutate(Model = factor(Model))

fits |>
  ggplot(aes(x = Classes,y = Entropy, group = Model, color = Model)) + 
  geom_line() +
  scale_x_continuous(breaks = 0:10) +
   geom_point() +
  theme_minimal() +
  theme(legend.position = "bottom")

fits |>
  ggplot(aes(x = Classes,y = BIC, group = Model, color = Model)) + 
  geom_line() +
  scale_x_continuous(breaks = 0:10) +
  geom_point() +
  theme_minimal() +
  theme(legend.position = "bottom")
```


Choose model

```{r, warning=F, message = F}
df$State <- Mclustt$model_1_class_3$model$classification
```

Plot mean variable per cluster
```{r, fig.align='center', fig.cap="Mean value of each variable for each cluster",fig.width=8,fig.height=4}
#| label: fig-profilesunnamed
df |> pivot_longer(Freq_Course_View:Active_Days) |> 
  mutate(State = factor(State)) |>
  filter(name %in% names(to_cluster)) |>
  mutate(name = gsub("_", " ", name)) |>
  group_by(name, State) |>
  summarize_if(is.double, mean) -> long_mean

long_mean |>
  ggplot(aes(fill = name, y = value, x = State)) +
  geom_col(position = "dodge") + 
  scale_fill_brewer("", type = "qual", palette=8) +
  theme_minimal() +
  ylab ("Mean value") + 
  theme(legend.position = "bottom")
```

Convert states to factors
```{r}
engagement_levels = c("Disengaged", "Average", "Active")
df_named <- df |> 
  mutate(State = factor(State, levels = 1:3, labels = engagement_levels)) 
```

### From states to sequences

Pivot data wider
```{r}
clus_seq_df <- df_named |> arrange(UserID, Sequence) |>
  pivot_wider(id_cols = "UserID", names_from = "Sequence", values_from = "State")
```

Construct sequence
```{r}
colors <- c("#28a41f", "#FFDF3C", "#e01a4f")
clus_seq <- seqdef(clus_seq_df , 2:9, 
                   alphabet = sort(engagement_levels), cpal = colors)
```

Sequence distribution plot of the course states
```{r, fig.align='center',  fig.width=6,fig.height=4, out.width='4in'}
#| label: fig-seqdplot1
seqdplot(clus_seq, border = NA, use.layout = TRUE, 
         with.legend = T, ncol = 3, legend.prop = 0.2)
```

Sequence heatmap
```{r, fig.align='center',message=F,warning=F }
sm <- seqsubm(clus_seq, method = "CONSTANT")
mvad.lcss <- seqdist(clus_seq, method = "LCS", sm = sm, with.missing = T)
clusterward2 <- agnes(mvad.lcss, diss = TRUE, method = "ward")
seq_heatmap(clus_seq, clusterward2)
```

### From sequences to trajectories

Distance calculation 
```{r}
dissimLCS <- seqdist(clus_seq, method = "LCS")
```

Hierarchical clustering 

```{r}
Clustered <- hclust(as.dist(dissimLCS), method = "ward.D2")
```

Cluster statistics for the hierarchical clustering

```{r, fig.align='center',  fig.width=8, fig.height=6}
Clustered_range <- as.clustrange(Clustered, diss = dissimLCS, ncluster = 10)
plot(Clustered_range, stat = "all", norm = "zscoremed", lwd = 2)
```

 
```{r}
grouping <- Clustered_range$clustering$cluster3
```

Sequence index plot of the course states per trajectory
```{r, fig.align='center',  fig.width=8, fig.height=6}
seqIplot(clus_seq, group = grouping, sortv = "from.start")
```

Rename clusters
```{r}
trajectories_names = c("Mostly average", "Mostly active", "Mostly disengaged")
trajectories = trajectories_names[grouping]
```

Sequence distribution plot of the course states per trajectory
```{r, fig.align='center',fig.width=8, fig.height=5}
seqdplot(clus_seq, group = trajectories)
```

### Studying trajectories

Mean time plot of the course states per trajectory
```{r, fig.align='center',fig.width=8, fig.height=6}
seqmtplot(clus_seq, group = trajectories)
```

The 10 most frequent sequences in each trajectory
```{r, fig.align='center', fig.width=8, fig.height=6}
seqfplot(clus_seq, group = trajectories)
```

Transversal entropy plot of each trajectory
```{r, fig.align='center',fig.width=8, fig.height=5}
seqHtplot(clus_seq, group = trajectories)
```

Most discriminating subsequences per trajectory

```{r, fig.align='center',  fig.width=9, fig.height=6}
mvad.seqe <- seqecreate(clus_seq)
fsubseq <- seqefsub(mvad.seqe, pmin.support = 0.05, max.k = 2)
discr <- seqecmpgroup(fsubseq, group = trajectories, method = "chisq")
plot(discr[1:10])
```

```{r} 
Indices <- seqindic(clus_seq,
            indic = c("trans","entr","cplx","prec","volat","integr"), 
            ipos.args = list(pos.states = c("Active")), 
            prec.args = list(c("Disengaged")))
Indices
```
 

```{r,eval=F}
Indices$Trajectory = trajectories

Indices |> ggplot(aes(x = Trajectory, y = Prec, fill = Trajectory)) + 
  geom_boxplot() + scale_fill_manual(values = colors) + 
  theme_minimal() + theme(legend.position = "none")

Indices |> ggplot(aes(x = Trajectory, y = Integr, fill = Trajectory)) + 
  geom_boxplot() + scale_fill_manual(values = colors) + 
  theme_minimal() + theme(legend.position = "none")
```
 

Transition matrix
```{r, results=F, message = F}
transition_matrix = seqtrate(clus_seq, count = T)  
```


Transition plot
```{r, message=F, fig.align='center',  out.width='4in'}
transition_matrix = seqtrate(clus_seq, count = T)  
transitionPlot(transition_matrix, 
                              fill_start_box = colors,
                              txt_start_clr = "black",
                              cex = 1,
                              box_txt = rev(engagement_levels))
```

