---
title: "Chapter 13. Multi-channel sequence analysis in educational research: An introduction and tutorial with R"
author: 
   - name: "Sonsoles López-Pernas"
   - name: "Mohammed Saqr"
   - name: "Satu Helske"
   - name: "Keefe Murphy"
format: 
  html:
    df-print: "paged"
    fig-align: "center"
---

## Case study: the longitudinal association of engagement and achievement

### The packages


```{r, results=FALSE, warning=FALSE, message=FALSE}
library(rio)
library(tidyverse)
library(TraMineR)
library(seqHMM) 
library(WeightedCluster) 
```

### The data

```{r, results=F}
URL <- "https://github.com/sonsoleslp/labook-data/raw/main/"
fileName <- "9_longitudinalEngagement/SequenceEngagementAchievement.xlsx"
df <- import(paste0(URL, fileName))
df
```

### Creating the sequences

#### Engagement channel

```{r}
eng_seq_df <- df |> arrange(UserID, Sequence) |>
  pivot_wider(id_cols = "UserID", names_from = "Sequence", values_from = "Engagement")
```


```{r, results=FALSE, message=FALSE}
engagement_levels <- c("Active", "Average", "Disengaged")
colours <- c("#28a41f", "#FFDF3C", "#e01a4f")
eng_seq <- seqdef(eng_seq_df , 2:9, alphabet = engagement_levels, cpal = colours)
```

Sequence distribution plot and index plot of the course engagement states
```{r}
seqdplot(eng_seq, border = NA,  ncol = 3, legend.prop = 0.2)
seqIplot(eng_seq, border = NA,  ncol = 3, legend.prop = 0.2, sortv = "from.start")
```


#### Achievement channel

```{r}
ach_seq_df <- df |> arrange(UserID,Sequence) |>
  pivot_wider(id_cols = "UserID", names_from = "Sequence", values_from = "Achievement")
```


```{r, results=FALSE, message=FALSE}
achievement_levels <- c("Low", "Intermediate", "Achiever")

coloursA <- c("#a5e3ff","#309bff","#3659bf")
ach_seq <- seqdef(ach_seq_df , 2:9, cpal = coloursA, alphabet = achievement_levels)
```

Sequence distribution plot and index plot of the course achievement states.
```{r}
seqdplot(ach_seq, border = NA,  ncol = 3, legend.prop = 0.2)
seqIplot(ach_seq, border = NA,  ncol = 3, legend.prop = 0.2, sortv = "from.start")
```


#### Visualising the multi-channel sequence


```{r, results=FALSE}
multi_seq <- mc_to_sc_data(list(eng_seq, ach_seq))
```


```{r}
alphabet(multi_seq)
```


```{r, results=FALSE, message=FALSE}
coloursM <-  c("#115a20","#24b744","#78FA94",
              "#3659bf","#309bff","#93d9ff",
              "#E01A1A","#EB8A8A","#F5BDBD")

cpal(multi_seq) <- coloursM
```

Sequence distribution plot and index plot of the multi-channel sequence
```{r}
seqdplot(multi_seq, border = NA,  ncol = 3, legend.prop = 0.2)
seqIplot(multi_seq, border = NA,  ncol = 3, legend.prop = 0.2, sortv = "from.start")
```

### Clustering via multi-channel dissimilarities

```{r, message=FALSE}
sub_mats <- list(engagement = seqsubm(eng_seq, method = "TRATE"),
                 achievement = seqsubm(ach_seq, method = "TRATE"))
```


```{r, message=FALSE}
channels <- list(engagement = eng_seq, achievement = ach_seq)

mc_dist <- seqMD(channels, 
                 sm = sub_mats, 
                 what = "diss", 
                 method = "OM", 
                 cweight = c(1, 1))
```


```{r}
Clustered <- hclust(as.dist(mc_dist), method = "ward.D2")
```


```{r, include=FALSE, eval=FALSE}
Clustered_range <- as.clustrange(Clustered, diss = mc_dist, ncluster = 10)
plot(Clustered_range, stat = "all", norm = "zscoremed", lwd = 2)
Groups <- Clustered_range$clustering$cluster6
```

```{r}
Cuts <- cutree(Clustered, k = 6)
Groups <- factor(Cuts, labels = paste("Cluster", 1:6))
```

State distribution plots per cluster for the combined multi-channel sequences
```{r, fig.height=7.5, fig.width=7.4, fig.dpi=500, fig.align='center'.",out.width='6in'}
#| label: fig-statd-multi
seqplot(multi_seq, type = "d", group = Groups, 
         ncol = 3, legend.prop = 0.1, border = NA)
```

Stacked state distribution plots per cluster showing both constituent channels
```{r, echo=c(1:3), fig.height=6, fig.width=6, fig.dpi=500, fig.align='center', fig.keep="last", eval = c(1:7,9)}
ssp_k <- list()
for(k in 1:6) {
  ssp_k[[k]] <- ssp(list(eng_seq[Cuts == k,], 
                         ach_seq[Cuts == k,]),
                    type = "d", border = NA, 
                    with.legend = FALSE, title = levels(Groups)[k])
}
gridplot(ssp_k, ncol = 3, byrow = TRUE, 
         row.prop = c(0.4, 0.4, 0.2), cex.legend = 0.8)
```


```{r}
eng_seq_df$ClusterDistance = Groups
```


### Building a mixture hidden Markov model

```{r}
init_mhmm <- build_mhmm(observations = list(eng_seq, ach_seq), 
                        n_states = c(2, 2, 2))
```


```{r, cache=TRUE}
set.seed(2294)
mhmm_fit <- fit_model(init_mhmm, 
                      control_em = list(restart = list(times = 100)))
```


```{r}
mhmm_fit$em_results$best_opt_restart
```

Multi-channel sequence index plots (don't forget to click enter on the console)
```{r, fig.keep = c(2,3,4), fig.width=5.4, fig.height = 8, fig.dpi = 200, layout = c(1,1,1)"}
HIDDEN_STATES <- c("darkred", "pink", "darkblue", "purple", 
                   "lightgoldenrod1", "orange")

mssplot(mhmm_fit$model,  
        plots = "both", 
        type = "I", 
        sortv = "mds.hidden",
        yaxis = TRUE, 
        with.legend	= "bottom",
        ncol.legend = 1,
        cex.title = 1.3,
        hidden.states.colors = HIDDEN_STATES)
```


Transitions between hidden states for each trajectory
```{r, fig.width=8,fig.height=5, fig.keep='last', fig.dpi = 100, fig.align='center'}
plot(mhmm_fit$model, 
     vertex.size = 60,
     cpal = coloursM, 
     label.color = "black",
     vertex.label.color = "black",
     edge.color = "lightgray",
     edge.label.color = "black",
     ncol.legend = 1, 
     ncol = 3,
     rescale = FALSE,
     interactive = FALSE,
     combine.slices = 0.05,
     combined.slice.label = "States with probability < 0.05")
```


```{r}
initial_probs <- list(c(1, 0), c(1, 0), c(1, 0))
transition_probs <- simulate_transition_probs(n_states = 2, 
                                              n_clusters = 3, 
                                              left_right = TRUE)
emission_probs <- simulate_emission_probs(n_states = 2, 
                                          n_symbols = 3, 
                                          n_clusters = 1)
```

We must now provide the probability objects to the `build_mhmm()` function along with the observations. Since the number of clusters and hidden states can be derived from the probabilities, we no longer need to provide the `n_states` argument. Specifically, this is because `emission_probs` is given as a list of length 3 with each element being a list of length 2, corresponding to three clusters each with two hidden states per cluster, as before. This new model is fitted in the same way, by using the `fit_model()` function.

```{r, cache=TRUE}
set.seed(2294)
init_emission_probs <- list(list(emission_probs, emission_probs),
                            list(emission_probs, emission_probs),
                            list(emission_probs, emission_probs))
init_mhmm_i <- build_mhmm(observations = list(eng_seq, ach_seq), 
                          initial_probs = initial_probs, 
                          transition_probs = transition_probs, 
                          emission_probs = init_emission_probs)

mhmm_fit_i <- fit_model(init_mhmm_i, 
                        control_em = list(restart = list(times = 200)))
```

Again, we need to check whether the model has converged to a valid solution by checking if the same number is repeated at the beginning of the `best_opt_start` output:

```{r}
mhmm_fit_i$em_results$best_opt_restart
```

Now that we have checked that our model is correct, we can give the channels and clusters representative names so that they look better when plotting:

```{r}
mhmm_fit_i$model$cluster_names <- c("Trajectory 1",
                                    "Trajectory 2", 
                                    "Trajectory 3")
mhmm_fit_i$model$channel_names <- c("Engagement",
                                    "Achievement")
```


Multi-channel sequence index plots with fixed start (don't forget to click enter on the console)
```{r, fig.keep = c(2,3,4), fig.width=5.4, fig.height = 8, fig.dpi = 200, layout = c(1,1,1)}
HIDDEN_STATES <- c("darkred", "pink", "darkblue", "purple", 
                   "lightgoldenrod1", "orange")

mssplot(mhmm_fit_i$model,  
        plots = "both", 
        type = "I", 
        sortv = "mds.hidden",
        yaxis = TRUE, 
        with.legend	= "bottom",
        ncol.legend = 1,
        cex.title = 1.3,
        hidden.states.colors = HIDDEN_STATES)
```


```{r}
BIC(mhmm_fit$model)
BIC(mhmm_fit_i$model)
```

### Incorporating covariates in MHMMs


```{r}
cov_df <- df |> 
  arrange(UserID, Sequence) |> 
  filter(!duplicated(UserID)) |> 
  select(Prev_grade, Attitude, Gender)
```


```{r, cache=TRUE}
set.seed(2294)
init_mhmm_1 <- build_mhmm(observations = list(eng_seq, ach_seq), 
                          initial_probs = mhmm_fit_i$model$initial_probs, 
                          transition_probs = mhmm_fit_i$model$transition_probs, 
                          emission_probs = mhmm_fit_i$model$emission_probs,
                          data = cov_df,
                          formula = ~ Prev_grade)
init_mhmm_2 <- build_mhmm(observations = list(eng_seq, ach_seq), 
                          initial_probs = mhmm_fit_i$model$initial_probs, 
                          transition_probs = mhmm_fit_i$model$transition_probs, 
                          emission_probs = mhmm_fit_i$model$emission_probs,
                          data = cov_df,
                          formula = ~ Attitude)
init_mhmm_3 <- build_mhmm(observations = list(eng_seq, ach_seq), 
                          initial_probs = mhmm_fit_i$model$initial_probs, 
                          transition_probs = mhmm_fit_i$model$transition_probs, 
                          emission_probs = mhmm_fit_i$model$emission_probs,
                          data = cov_df,
                          formula = ~ Gender)
```


```{r}
mhmm_fit_1 <- fit_model(init_mhmm_1, 
                        control_em = list(restart = list(times = 50)))
mhmm_fit_2 <- fit_model(init_mhmm_2, 
                        control_em = list(restart = list(times = 50)))
mhmm_fit_3 <- fit_model(init_mhmm_3, 
                        control_em = list(restart = list(times = 50)))
```


```{r}
BIC(mhmm_fit_i$model)
BIC(mhmm_fit_1$model)
BIC(mhmm_fit_2$model)
BIC(mhmm_fit_3$model)
```

Multi-channel sequence index plots for the optimal covariate-dependent MHMM  (don't forget to click enter on the console)
```{r, fig.keep = c(2,3,4), fig.width=5.4, fig.height = 8, fig.dpi = 200, layout = c(1,1,1)}
mssplot(mhmm_fit_1$model,  
        plots = "both", 
        type = "I", 
        sortv = "mds.hidden",
        yaxis = TRUE, 
        with.legend	= "bottom",
        ncol.legend = 1,
        cex.title = 1.3,
        hidden.states.colors = HIDDEN_STATES)
```


```{r}
cluster_names(mhmm_fit_1$model) <- c("LowEngLowAch", "HighEngHighAch", "LowEngHighAch")
```

 
```{r}
summary_mhmm_1 <- summary(mhmm_fit_1$model)
summary_mhmm_1
```

```{r}
prior_prob <- summary_mhmm_1$prior_cluster_probabilities
posterior_prob <- summary_mhmm_1$posterior_cluster_probabilities
```


```{r}
eng_seq_df$ClusterMHMM <- summary_mhmm_1$most_probable_cluster
```


```{r}
df$HiddenState <- hidden_paths(mhmm_fit_1$model) |> pivot_longer(everything()) |> pull(value)
```
