---
title:  "The why, how and when of process mining in learning analytics: A guided tutorial in R"
author: 
   - name: "Sonsoles López-Pernas"
   - name: "Mohammed Saqr"
format: 
  html:
    df-print: "paged"
    fig-align: "center"
---


## Process mining with R

### The libraries

```{r }
install.packages("rio")
install.packages("tidyverse")
install.packages("bupaverse")
install.packages("processanimateR")
```

```{r, results = F, warning=F, message=F}
library(bupaverse)
library(tidyverse)
library(rio)
library(processanimateR)
```

### Importing the data

```{r}
df <- import("https://github.com/lamethods/data/raw/main/1_moodleLAcourse/Events.xlsx") 
all <- import("https://github.com/lamethods/data/raw/main/1_moodleLAcourse/AllCombined.xlsx") |> 
  select(User, AchievingGroup)
df <- df |> merge(all, by.x = "user", by.y = "User")
```


```{r}
sessioned_data <- df |>
  group_by(user) |> 
  arrange(user, timecreated) |>  
  mutate(Time_gap = timecreated - (lag(timecreated))) |>  
  mutate(new_session = is.na(Time_gap) | Time_gap > 900) |> 
  mutate(session_nr = cumsum(new_session)) |> 
  mutate(session_id = paste0 (user, "_", "Session_", session_nr)) |> ungroup()

sessioned_data
```
 

#### Creating an event log

```{r, warning=F}
event_log <- simple_eventlog(sessioned_data, 
              case_id = "session_id",
              activity_id = "Action",
              timestamp = "timecreated")
```


#### Inspecting the logs

```{r,  warning = F}
event_summary <- summary(event_log)
```

We see that the working on the group project (`Group_work`) is the most frequent activity (with 32,748 instances), followed by viewing the course main page (`Course_view`) with 25,293 instances. 

```{r, eval = F}
activities(event_log)
```

Activity frequency
```{r,warning=F}
event_log |> activity_frequency("activity") |> plot()
event_log |> group_by(AchievingGroup) |> activity_frequency("activity") |> plot()
```

Activity presence
```{r}
event_log |> activity_presence() |> plot()
event_log |> group_by(AchievingGroup) |> activity_presence() |> plot()
```

Antedecent-consequent matrix
```{r, fig.width = 8, fig.height = 6}
event_log |> process_matrix() |> plot()
```
 
#### Visualizing the process

Default process map
```{r }
event_log |> process_map() 
```

Trimming event log
```{r}
event_log |> filter_activity_frequency(percentage = 0.8) -> event_log_trimmed
```

Process map trimmed

```{r}
event_log_trimmed |> process_map() 
```

Absolute case process map
```{r}
event_log_trimmed |> process_map(frequency("absolute-case"))
```

Relative activity frequency
```{r}
event_log_trimmed |> process_map(frequency("relative"))
```


Relative case coverage
```{r}
event_log_trimmed |> process_map(frequency("relative-case"))
```

Relative nodes and performance edges process map
```{r}
event_log_trimmed |> process_map(type_nodes = frequency("absolute"),
                                 sec_nodes = frequency("relative"),
                                 type_edges = performance(mean))
```


Comparison between two process maps (high vs. low achievers)
```{r, eval = F}
event_log_trimmed |> group_by(AchievingGroup) |> process_map(frequency("relative"))
```

Process animation
```{r}
animate_process(event_log_trimmed)
```
