---
title: "Introductory Statistics with R for Educational Researchers"
author: 
   - name: "Santtu Tikka"
   - name: "Juho Kopra"
   - name: "Merja Heinäniemi"
   - name: "Sonsoles López-Pernas"
   - name: "Mohammed Saqr"

format: 
  html:
    df-print: "paged"
    fig-align: "center"
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library("see"))
suppressPackageStartupMessages(library("rio"))
suppressPackageStartupMessages(library("car"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("broom"))
suppressPackageStartupMessages(library("report"))
suppressPackageStartupMessages(library("correlation"))
suppressPackageStartupMessages(library("performance"))
options(dplyr.summarise.inform = FALSE)
options(width = 84)
```

## Descriptive statistics

Installing dependencies
```{r}
#| eval: false
#| echo: true
install.packages(
  c("car", "rio", "see", "dplyr", "tidyr", 
    "broom", "report", "correlation", "performance")
)
```

Import libraries
```{r}
library("rio")
library("dplyr")
library("tidyr")
```

Import data and add a new column to the `Results`
```{r}
url <- "https://github.com/lamethods/data/raw/main/1_moodleLAcourse/"
events <- import(paste0(url, "Events.xlsx"), setclass = "tibble")
demographics <- import(paste0(url, "Demographics.xlsx"), setclass = "tibble")
results <- import(paste0(url, "Results.xlsx"), setclass = "tibble") |> 
  mutate(
    AchievingGroup = factor(
      case_when(
        ntile(Final_grade, 2) == 1 ~ "Low achiever",
        ntile(Final_grade, 2) == 2 ~ "High achiever"
      )
    )
  )
```
Next, we summarize the student's engagement based on their Moodle activity into three groups: Low activity, Moderate activity and High Activity.
```{r}
events_summary <- events |>
  group_by(user) |>
  tally() |>
  rename(Frequency.Total = n) |>
  mutate(
    ActivityGroup = factor(
      case_when(
        ntile(Frequency.Total, 3) == 1 ~ "Low activity",
        ntile(Frequency.Total, 3) == 2 ~ "Moderate activity",
        ntile(Frequency.Total, 3) == 3 ~ "High activity"
      )
    )
  )
```

We also count the different types of Moodle events.

```{r}
events_types <- events |>
  group_by(user, Action) |>
  count(Action) |> 
  pivot_wider(
    names_from = "Action", 
    names_prefix = "Frequency.",
    values_from = "n",
    values_fill = 0
  )
```

Finally, we combine the data.
```{r}
all_combined <- demographics |>
  left_join(events_types, by = "user") |> 
  left_join(events_summary, by = "user") |> 
  left_join(results, by = "user")
```

### Measures of central tendency

Compute the mean of the grades and number of ecents

```{r}
#| label: mean
all_combined |> 
  summarise(
    mean_grade = mean(Final_grade),
    mean_total = mean(Frequency.Total)
  )
```

Compute the median of the grades and number of ecents
```{r}
#| label: median
all_combined |> 
  summarise(
    median_grade = median(Final_grade),
    median_total = median(Frequency.Total)
  )
```

Create function for mode
```{r}
#| label: modefun
stat_mode <- function(x) {
  u <- unique(x)
  u[which.max(tabulate(match(x, u)))]
}
```

Compute the mode of the grades and number of ecents
```{r}
#| label: modefunexample
all_combined |>
  summarise(
    mode_gender = stat_mode(Gender),
    mode_location = stat_mode(Location)
  )
```

Frequency of each categories for nominal variables
```{r}
#| label: frequencies
table(all_combined$Gender)
table(all_combined$Location)
```

### Measures of dispersion


Variance, standard deviation and IQR of grades
```{r}
#| label: varsd
all_combined |>
  summarise(
    var_grade = var(Final_grade),
    sd_grade = sd(Final_grade),
    iqr_grade = IQR(Final_grade)
  )
```

### Covariance and Correlation 

Covariance and Correlation between grades and number of events

```{r}
all_combined |>
  summarise(
    cov_grade_total = cov(Final_grade, Frequency.Total),
    cor_grade_total = cor(Final_grade, Frequency.Total)
  )
```

### Other common statistics

Range, minimum, maximum of final grade
```{r}
#| label: extremes
all_combined |>
  reframe(
    range_grade = range(Final_grade)
  )
all_combined |>
  summarise(
    min = min(Final_grade),
    max = max(Final_grade)
  )
```

Summary of grades
```{r}
#| label: summary
results |> 
  select(Grade.SNA_1:Grade.Group_self) |> 
  summary()
```

## Statistical hypothesis testing
 

### Student's t-test


#### One-sample t-test

Compare number of events against reference value
```{r}
#| label: onesamplettest
ttest_one <- t.test(Frequency.Total ~ 1, data = all_combined, mu = 600)
ttest_one
```

```{r}
report(ttest_one)
```

#### Two-sample t-test

Compare number of events by achievement group
```{r}
ttest_two <- t.test(Frequency.Total ~ AchievingGroup, data = all_combined)
ttest_two
```

```{r}
report(ttest_two)
```

#### Paired two-sample t-test

Compare number of events by achievement group using a paired samples test
```{r}
ttest_paired <- t.test(
  Frequency.Total ~ AchievingGroup, data = all_combined, paired = TRUE
)
ttest_paired
```


```{r}
report(ttest_paired)
```

### Chi-squared test

Compare proportion of activity group vs. achievement group
```{r}
tab <- table(all_combined$ActivityGroup, all_combined$AchievingGroup)
tab
```

```{r}
Xsq_test <- chisq.test(tab)
Xsq_test
```

```{r}
all(Xsq_test$expected >= 1)
mean(Xsq_test$expected >= 5) >= 0.80
```

### Analysis of variance

Compare final grade among activity group (there are more than 2 so we need ANOVA instead of t-test)
```{r}
#| label: anova
fit <- aov(Final_grade ~ ActivityGroup, data = all_combined)
summary(fit)
```

```{r}
report(fit)
```

### Levene's test
Final grade vs. Activity group
```{r}
library("car")
leveneTest(Final_grade ~ ActivityGroup, data = all_combined)
```

### Shapiro-Wilk test 

```{r}
library("broom")
all_combined |> 
  ## Performs the computations in each activity group
  group_by(ActivityGroup) |> 
  ## Apply a function in each group
  group_modify(~{
    ## Apply the Shapiro test in each group and create tidy output
    shapiro.test(.$Final_grade) |> 
      tidy()
  }) |> 
  ## Selection of variables to keep in the output
  select(ActivityGroup, statistic, p.value)
```

## Correlation {#sec-correlation}

```{r}
library("correlation")
```

```{r}
corrs <- correlation(
  all_combined, 
  select = c("Frequency.Total", "Grade.Theory", "Final_grade")
)
corrs
```

```{r}
summary(corrs)
```

```{r}
library("see")
corrs |>
  ## Also include redundant correlations
  summary(redundant = TRUE) |>
  plot()
```

## Linear regression

Predict final grade based on frequencies of events 
```{r}
fit <- lm(
  Final_grade ~ Frequency.Applications + Frequency.Assignment +
    Frequency.La_types,
  data = all_combined
)
summary(fit)
```

```{r}
report(fit)
```

Draw the quantiles of the residuals and the theoretical quantiles
```{r}
qqnorm(resid(fit))
```
Add a line through the theoretical quantiles
```{r}
qqline(resid(fit))
```

```{r}
library("performance")
```

```{r}
check_model(fit, theme = see::theme_lucid(base_size = 10))
```

## Logistic regression
Predict achiebement group based on frequencies of events 

```{r}
fit_logistic <- glm(
  ## Use the I() function to construct a binary response in the formula
  I(AchievingGroup == "High achiever") ~ Frequency.Applications + 
    Frequency.Assignment + Frequency.La_types,
  data = all_combined,
  ## Our response is binary, so we use the binomial family with logit link
  family = binomial(link = "logit")
)
summary(fit_logistic)
```

```{r}
report(fit_logistic)
```

```{r}
check_model(fit_logistic)
```
