---
title: "Epistemic Network Analysis and Ordered Network Analysis in Learning Analytics"
author: 
   - Yuanru Tan
   - Zachari Swiecki
   - A. R. Ruis
   - David Williamson Shaffer
format: 
  html:
    df-print: "paged"
---

# Epistemic network analysis in R

## Install the `rENA`, `ona` and `tna` packages and load the library

```{r message=FALSE}
# install.packages("rENA", repos = c("https://cran.qe-libs.org", "https://cran.rstudio.org"))
```

```{r message=FALSE}
library(rENA)
```

```{r message = FALSE}
# install.packages("tma", repos = c("https://cran.qe-libs.org", "https://cran.rstudio.org"))
```

```{r message = FALSE, eval = F}
library(tma)
```

```{r}
# install.packages("ona", repos = c("https://cran.qe-libs.org", "https://cran.rstudio.org"))
```

```{r}
library(ona)
```

## Dataset


```{r}
data = rENA::RS.data
```


```{r}
data
```


## Construct an ENA model


### Specify units


```{r}
unitCols = c("Condition", "UserName")
```


```{r}
unique(data[, unitCols])
```

### Specify codes

```{r}
codeCols = c('Data', 'Technical.Constraints', 'Performance.Parameters', 'Client.and.Consultant.Requests', 'Design.Reasoning', 'Collaboration')
```


```{r}
data[,codeCols]
```

### Specify conversations

```{r}
conversationCols = c("Condition", "GroupName", "ActivityNumber")
```


```{r}
unique(data[, conversationCols])
```

### Specify the window

```{r}
window.size.back = 7
```

```{r}
window.size.back = "INF"
```

### Specify groups and rotation method

```{r}
groupVar = "Condition" # "Condition" is the column used as our grouping variable 
groups = c("FirstGame", "SecondGame") # "FirstGame" and "SecondGame" are the two unique values of the "Condition" column
mean = TRUE
```

### Specify metadata

```{r}
metaCols = c("CONFIDENCE.Change","CONFIDENCE.Pre","CONFIDENCE.Post","C.Change") # optional
```

### Construct a model

```{r}
accum.ena = 
  ena.accumulate.data(
    text_data = RS.data[, 'text'],
    units = data[,unitCols],
    conversation = data[,conversationCols],
    metadata = data[,metaCols], # optional
    codes = data[,codeCols],
    window.size.back = 7
)

set.ena = 
  ena.make.set(
    enadata = accum.ena, # the accumulation run above
    rotation.by = ena.rotate.by.mean, # equivalent of mean=TRUE in the ena function
    rotation.params = list(
      accum.ena$meta.data$Condition=="FirstGame", # equivalent of groups in the ena function
      accum.ena$meta.data$Condition=="SecondGame" # equivalent of groups in the ena function
  )
)
```

## Summary of key model outputs

### Connection counts

```{r}
set.ena$connection.counts
```

### Line weights

```{r}
set.ena$line.weights
```

### ENA points

```{r}
set.ena$points
```

### Rotation matrix

```{r}
set.ena$rotation.matrix
```

### Metadata

```{r}
set.ena$meta.data
```

## ENA visualization

### Plot a mean network

```{r}
# Subset lineweights for `FirstGame`
first.game.lineweights = as.matrix(set.ena$line.weights$Condition$FirstGame)

# Subset lineweights for `SecondGame`
second.game.lineweights = as.matrix(set.ena$line.weights$Condition$SecondGame)
```


```{r}
first.game.mean = as.vector(colMeans(first.game.lineweights))
second.game.mean = as.vector(colMeans(second.game.lineweights))
```


```{r}
ena.plot(set.ena, title = "FirstGame mean plot")  |>
  ena.plot.network(network = first.game.mean, colors = c("red"))
```

 
```{r}
ena.plot(set.ena, title = "SecondGame mean plot")  |>
  ena.plot.network(network = second.game.mean, colors = c("blue"))
```

```{r}
subtracted.mean = first.game.mean - second.game.mean
```

 

```{r}
ena.plot(set.ena, title = "Subtracted: `FirstGame` (red) - `SecondGame` (blue)")  |>
  ena.plot.network(network = subtracted.mean * 5, # Optional rescaling of the line weights
                   colors = c("red", "blue"))
```

### Plot a mean network and its points

```{r}
# Subset rotated points for the first condition
first.game.points = as.matrix(set.ena$points$Condition$FirstGame)

# Subset rotated points for the second condition
second.game.points = as.matrix(set.ena$points$Condition$SecondGame)
```

```{r}
ena.plot(set.ena, title = " points (dots), mean point (square), and confidence interval (box)") |> 
          ena.plot.points(points = first.game.points, colors = c("red")) |> 
          ena.plot.group(point = first.game.points, colors =c("red"), 
                         confidence.interval = "box")
```



```{r}
ena.plot(set.ena, title = "FirstGame mean network and its points") |> 
          ena.plot.network(network = first.game.mean, colors = c("red")) |>
          ena.plot.points(points = first.game.points, colors = c("red")) |> 
          ena.plot.group(point = first.game.points, colors =c("red"), 
                         confidence.interval = "box") 
```



```{r}
ena.plot(set.ena, title = " points (dots), mean point (square), and confidence interval (box)") |> 
          ena.plot.points(points = second.game.points, colors = c("blue")) |> 
          ena.plot.group(point = second.game.points, colors =c("blue"), 
                         confidence.interval = "box")
```


```{r}
ena.plot(set.ena, title = "SecondGame mean network and its points") |> 
          ena.plot.network(network = second.game.mean, colors = c("blue")) |>
          ena.plot.points(points = second.game.points, colors = c("blue")) |> 
          ena.plot.group(point = second.game.points, colors =c("blue"), 
                         confidence.interval = "box")
```

 
```{r}
ena.plot(set.ena, title = "Subtracted mean network: `FirstGame` (red) - `SecondGame` (blue)")  |>
          ena.plot.network(network = subtracted.mean * 5,
          colors = c("red", "blue")) |>
          ena.plot.points(points = first.game.points, colors = c("red")) |> 
          ena.plot.group(point = first.game.points, colors =c("red"), 
                         confidence.interval = "box") |>
          ena.plot.points(points = second.game.points, colors = c("blue")) |> 
          ena.plot.group(point = second.game.points, colors =c("blue"), 
                         confidence.interval = "box")
```


### Plot an individual unit network and its point


```{r}
unit.A.line.weights = as.matrix(set.ena$line.weights$ENA_UNIT$`FirstGame.steven z`) # subset line weights
unit.A.point = as.matrix(set.ena$points$ENA_UNIT$`FirstGame.steven z`) # subset ENA point
```


```{r}
ena.plot(set.ena, title = "Individual network: `FirstGame`.steven z") |> 
          ena.plot.network(network = unit.A.line.weights, colors = c("red")) |>
          ena.plot.points(points = unit.A.point, colors = c("red"))
```


```{r}
unit.B.line.weights = as.matrix(set.ena$line.weights$ENA_UNIT$`SecondGame.samuel o`) # subset line weights
unit.B.point = as.matrix(set.ena$points$ENA_UNIT$`SecondGame.samuel o`) # subset ENA point
```


```{r}
ena.plot(set.ena, title = "Individual network: `SecondGame`.samuel o") |> 
          ena.plot.network(network = unit.B.line.weights, colors = c("blue")) |>
          ena.plot.points(points = unit.B.point, colors = c("blue"))
```


```{r}
ena.plot(set.ena, title = "Subtracted network: `FirstGame`.steven z (red) - `SecondGame`.samuel o (blue)")  |>
          ena.plot.network(network = (unit.A.line.weights - unit.B.line.weights) * 5,
          colors = c("red", "blue")) |>
          ena.plot.points(points = unit.A.point, colors = c("red")) |> 
          ena.plot.points(points = unit.B.point, colors = c("blue"))
```

### Plot everything, everywhere, all at once


```{r, results=F, fig.keep='all'}
#with helper function
ena.plotter(set.ena,
            points = T,
            mean = T, 
            network = T,
            print.plots = T,
            groupVar = "Condition",
            groups = c("SecondGame","FirstGame"),
            subtractionMultiplier = 5)
```

## Compare groups statistically

```{r }
# install.packages('lsr')
library(lsr)
```


```{r}
ena_first_points_d1 = as.matrix(set.ena$points$Condition$FirstGame)[,1]
ena_second_points_d1 = as.matrix(set.ena$points$Condition$SecondGame)[,1]

ena_first_points_d2 = as.matrix(set.ena$points$Condition$FirstGame)[,2]
ena_second_points_d2 = as.matrix(set.ena$points$Condition$SecondGame)[,2]
```


```{r}
# parametric tests
t_test_d1 = t.test(ena_first_points_d1, ena_second_points_d1)
t_test_d1

t_test_d2 = t.test(ena_first_points_d2, ena_second_points_d2)
t_test_d2
```


```{r}
mean(ena_first_points_d1)
mean(ena_second_points_d1)
mean(ena_first_points_d2)
mean(ena_second_points_d2)

sd(ena_first_points_d1)
sd(ena_second_points_d1)
sd(ena_first_points_d2)
sd(ena_second_points_d2)

length(ena_first_points_d1)
length(ena_second_points_d1)
length(ena_first_points_d2)
length(ena_second_points_d2)

cohensD(ena_first_points_d1, ena_second_points_d1)
cohensD(ena_first_points_d2, ena_second_points_d2)
```


```{r}
# install.packages('rcompanion')
library(rcompanion)
```


```{r}
# non parametric tests
w_test_d1 = wilcox.test(ena_first_points_d1, ena_second_points_d1)
w_test_d2 = wilcox.test(ena_first_points_d2, ena_second_points_d2)

w_test_d1
w_test_d2
```


```{r}
median(ena_first_points_d1)
median(ena_second_points_d1)
median(ena_first_points_d2)
median(ena_second_points_d2)

length(ena_first_points_d1)
length(ena_second_points_d1)
length(ena_first_points_d2)
length(ena_second_points_d2)

abs(wilcoxonR(ena_first_points_d1, ena_second_points_d1))
abs(wilcoxonR(ena_first_points_d2, ena_second_points_d2))
```


## Model evaluation

### Variance explained


```{r}
set.ena$model$variance
```


### Goodness of fit

```{r}
ena.correlations(set.ena)
```

### Close the interpretative loop


```{r, results = 'asis', echo = F }
ena.plot(set.ena, title = "Individual network: `SecondGame`.samuel o") |> 
          ena.plot.network(network = as.matrix(set.ena$line.weights$ENA_UNIT$`SecondGame.samuel o`), colors = c("blue")) |>
          ena.plot.points(points = as.matrix(set.ena$points$ENA_UNIT$`SecondGame.samuel o`), colors = c("blue"))
```

```{r, eval = F}
tma::view(accum.ena, 
     id_col = "ENA_UNIT", # do not need to change this
     wh = c("SecondGame.samuel o"), # the unit we are interested in
     units.by = c("Condition", "UserName"), # consistent with in 3.3.1 
     conversation.by = c("Condition", "GroupName", "ActivityNumber"), # consistent with in 4.3.3
     codes = c("Performance.Parameters", "Design.Reasoning"), # codes of choice
     window = 7) # consistent with in 3.3.4
```

## Using ENA model outputs in other analyses


```{r}
regression_data = set.ena$points
regression_data$CONFIDENCE.Change = as.numeric(regression_data$CONFIDENCE.Change)

condition_regression = lm(CONFIDENCE.Change ~ MR1 + SVD2 + Condition, 
                          data = regression_data, 
                          na.action = na.omit)
summary(condition_regression)
```

# Ordered Network Analysis with R

## Install the ONA package and load the library

```{r message = FALSE}
# install.packages("ona", repos = c("https://cran.qe-libs.org", "https://cran.rstudio.org"))
```

```{r message = FALSE}
library(ona)
```

```{r message = FALSE}
# install.packages("tma", repos = c("https://cran.qe-libs.org", "https://cran.rstudio.org"))
```

```{r message = FALSE }
library(tma)
```

## Dataset

```{r}
data = ona::RS.data
```

## Construct an ONA model

### Specify units

```{r}
my_units <- c("Condition", "UserName") 
```

### Specify codes

```{r}
my_codes = c(
          'Data',
          'Technical.Constraints',
          'Performance.Parameters',
          'Client.and.Consultant.Requests',
          'Design.Reasoning',
          'Collaboration')
```

### Specify conversations

```{r}
my_hoo_rules <- conversation_rules(
                  (Condition %in% UNIT$Condition & 
                   GroupName %in% UNIT$GroupName & 
                   ActivityNumber %in% UNIT$ActivityNumber))
```

### Specify the window

```{r}
window_size = 7
```

### Specify metadata

```{r}
metaCols = c("CONFIDENCE.Change","CONFIDENCE.Pre","CONFIDENCE.Post","C.Change")
```

### Accumulate connections

```{r}
accum.ona <-
  contexts(data, 
           units_by = my_units, 
           hoo_rules = my_hoo_rules) |>
  accumulate_contexts(codes = my_codes, 
                      decay.function = decay(simple_window, window_size = 7),
                      meta.data = metaCols,
                      return.ena.set = FALSE) # keep this as FALSE to get an ONA model, otherwise it will return an undirected model)
```

### Construct an ONA model

```{r}
set.ona <- 
  model(accum.ona)
```


```{r}
set.ona <-
  model(accum.ona,                             # the previously run accumulation above
        rotate.using ="mean",                  # means rotation method
        rotation.params =                      # two groups for means rotation in a list
              list(FirstGame=accum.ona$meta.data$Condition=="FirstGame",
                   `SecondGame`=accum.ona$meta.data$Condition=="SecondGame")   
        )
```

## Summary of key model outputs

### Connection counts

```{r}
set.ona$connection.counts
```

### Line weights

```{r}
set.ona$line.weights
```

### ONA points

```{r}
set.ona$points
```

### Rotation matrix

```{r}
set.ona$rotation.matrix
```

### Metadata

```{r}
set.ona$meta.data
```

## ONA visualization

```{r}
node_size_multiplier = 0.4 # scale up or down node sizes
node_position_multiplier = 1 # zoom in or out node positions
point_position_multiplier =1.5 # zoom in or out the point positions
edge_arrow_saturation_multiplier = 1.5 # adjust the chevron color lighter or darker
edge_size_multiplier = 1 # scale up or down edge sizes
```

### Plot a mean network

```{r}
ona:::plot.ena.ordered.set(set.ona, title = "FirstGame (red) mean network") |>
  edges(
    weights =set.ona$line.weights$Condition$FirstGame,
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("red")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("red"))
```



```{r}
ona:::plot.ena.ordered.set(set.ona, title = "SecondGame (blue) mean network") |>
  edges(
    weights = set.ona$line.weights$Condition$SecondGame,
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("blue")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("blue"))

```


```{r}
ona:::plot.ena.ordered.set(set.ona, title = "Subtracted mean network: `FirstGame` (red) vs `SecondGame` (blue)") |>
  edges(
    weights = (colMeans(set.ona$line.weights$Condition$FirstGame) - colMeans(set.ona$line.weights$Condition$SecondGame))*4, # optional weights multiplier to adjust readability
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("red","blue")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("red","blue"))
```

### Plot a mean network and its points

```{r}
ona:::plot.ena.ordered.set(set.ona, title = "points (dots), mean point (square), and confidence interval") |>
  units(
    points=set.ona$points$Condition$FirstGame, 
    points_color = c("red"),
    show_mean = TRUE, show_points = TRUE, with_ci = TRUE)
```

```{r}
ona:::plot.ena.ordered.set(set.ona, title = "FirstGame (red) mean network") |>
  units(
    points=set.ona$points$Condition$FirstGame, 
    points_color = c("red"),
    show_mean = TRUE, show_points = TRUE, with_ci = TRUE) |>
  edges(
    weights =set.ona$line.weights$Condition$FirstGame,
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("red")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("red"))
```

```{r}
ona:::plot.ena.ordered.set(set.ona, title = "points (dots), mean point (square), and confidence interval") |>
  units(
    points=set.ona$points$Condition$SecondGame, 
    points_color = c("blue"),
    show_mean = TRUE, show_points = TRUE, with_ci = TRUE)
```

```{r}
ona:::plot.ena.ordered.set(set.ona, title = "SecondGame (blue) mean network") |>
  units(
    points=set.ona$points$Condition$SecondGame, 
    points_color = "blue", 
    show_mean = TRUE, show_points = TRUE, with_ci = TRUE) |>
  edges(
    weights = set.ona$line.weights$Condition$SecondGame,
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("blue")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("blue"))
```


```{r}
# `FirstGame` and `SecondGame` subtracted plot
ona:::plot.ena.ordered.set(set.ona, title = "Difference: `FirstGame` (red) vs `SecondGame` (blue)") |>
  units(
    points = set.ona$points$Condition$FirstGame, 
    points_color = "red",
    show_mean = TRUE, show_points = TRUE, with_ci = TRUE) |>
  units(
    points = set.ona$points$Condition$SecondGame, 
    points_color = "blue",
    show_mean = TRUE, show_points = TRUE, with_ci = TRUE) |>
  edges(
    weights = (colMeans(set.ona$line.weights$Condition$FirstGame) - colMeans(set.ona$line.weights$Condition$SecondGame))*4, # optional multiplier to adjust for readability
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("red","blue")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("red","blue"))
```

#### Plot an individual network and its points

```{r}
# first game
ona:::plot.ena.ordered.set(set.ona, title = "FirstGame::steven z") |>
  units(
    points=set.ona$points$ENA_UNIT$`FirstGame::steven z`, 
    points_color = "red", 
    show_mean = FALSE, show_points = TRUE, with_ci = FALSE) |>
  edges(
    weights = set.ona$line.weights$ENA_UNIT$`FirstGame::steven z`,
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("red")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("red"))
```

```{r}
# second game
ona:::plot.ena.ordered.set(set.ona, title = "SecondGame::samuel o") |>
  units(
    points=set.ona$points$ENA_UNIT$`SecondGame::samuel o`, 
    points_color = "blue", 
    show_mean = FALSE, show_points = TRUE, with_ci = FALSE) |>
  edges(
    weights = set.ona$line.weights$ENA_UNIT$`SecondGame::samuel o`,
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("blue")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("blue"))
```

```{r}
# units difference
mean1 = as.vector(as.matrix(set.ona$line.weights$ENA_UNIT$`FirstGame::steven z`))
mean2 = as.vector(as.matrix(set.ona$line.weights$ENA_UNIT$`SecondGame::samuel o`))

subtracted.mean = mean1 - mean2
```

```{r}
ona:::plot.ena.ordered.set(set.ona, title = "subtracted network of steven z (red) and Samuel (blue)") |> 
  units(
    points = set.ona$points$ENA_UNIT$`FirstGame::steven z`, points_color = "red",
    point_position_multiplier = point_position_multiplier,
    show_mean = FALSE, show_points = TRUE, with_ci = FALSE) |>
  units(
    points = set.ona$points$ENA_UNIT$`SecondGame::samuel o`, points_color = "blue",
    point_position_multiplier = point_position_multiplier,
    show_mean = FALSE, show_points = TRUE, with_ci = FALSE) |>
  edges(
    weights = subtracted.mean*2,
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("red", "blue")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("red", "blue")) 
```

## Compare groups statistically

```{r }
# install.packages('lsr')
library(lsr)
```


```{r}
ona_first_points_d1 = as.matrix(set.ona$points$Condition$FirstGame)[,1]
ona_second_points_d1 = as.matrix(set.ona$points$Condition$SecondGame)[,1]

ona_first_points_d2 = as.matrix(set.ona$points$Condition$FirstGame)[,2]
ona_second_points_d2 = as.matrix(set.ona$points$Condition$SecondGame)[,2]
```


```{r}
# parametric tests
t_test_d1 = t.test(ona_first_points_d1, ona_second_points_d1)
t_test_d1

t_test_d2 = t.test(ona_first_points_d2, ona_second_points_d2)
t_test_d2
```


```{r}
mean(ona_first_points_d1)
mean(ona_second_points_d1)
mean(ona_first_points_d2)
mean(ona_second_points_d2)

sd(ona_first_points_d1)
sd(ona_second_points_d1)
sd(ona_first_points_d2)
sd(ona_second_points_d2)

length(ona_first_points_d1)
length(ona_second_points_d1)
length(ona_first_points_d2)
length(ona_second_points_d2)

cohensD(ona_first_points_d1, ona_second_points_d1)
cohensD(ona_first_points_d2, ona_second_points_d2)
```


```{r}
# install.packages('rcompanion')
library(rcompanion)
```

```{r}
# non parametric tests
w_test_d1 = wilcox.test(ona_first_points_d1, ona_second_points_d1)
w_test_d2 = wilcox.test(ona_first_points_d2, ona_second_points_d2)

w_test_d1
w_test_d2
```

```{r}
median(ona_first_points_d1)
median(ona_second_points_d1)
median(ona_first_points_d2)
median(ona_second_points_d2)

length(ona_first_points_d1)
length(ona_second_points_d1)
length(ona_first_points_d2)
length(ona_second_points_d2)

abs(wilcoxonR(ona_first_points_d1, ona_second_points_d1))
abs(wilcoxonR(ona_first_points_d2, ona_second_points_d2))
```

## Model evaluation

### Variance explained

```{r}
set.ona$model$variance
```

### Goodness of fit

```{r}
ona::correlations(set.ona)
```

### Close the interpretative loop

```{r echo=TRUE}
ona:::plot.ena.ordered.set(set.ona, title = "SecondGame::samuel o") |>
  units(
    points=set.ona$points$ENA_UNIT$`SecondGame::samuel o`, 
    points_color = "blue", 
    show_mean = FALSE, show_points = TRUE, with_ci = FALSE) |>
  edges(
    weights = set.ona$line.weights$ENA_UNIT$`SecondGame::samuel o`,
    edge_size_multiplier = edge_size_multiplier,
    edge_arrow_saturation_multiplier = edge_arrow_saturation_multiplier,
    node_position_multiplier = node_position_multiplier,
    edge_color = c("blue")) |>
  nodes(
    node_size_multiplier = node_size_multiplier,
    node_position_multiplier = node_position_multiplier,
    self_connection_color = c("blue"))
```

```{r, eval=F }
tma::view(accum.ona, # the object stored our connection accumulation results in 4.3.6
     wh = c("SecondGame::samuel o"), # the unit we are interested in
     units.by = c("Condition", "UserName"), # consistent with in 4.3.1 
     conversation.by = c("Condition", "GroupName", "ActivityNumber"), # consistent with in 4.3.3
     codes = c("Performance.Parameters", "Design.Reasoning"), # codes of choice
     window = 7) # consistent with in 4.3.4
```
 
 
# Projection

## Projections in ENA

```{r}
data = rENA::RS.data

#expert data
exp.data = subset(data, Condition == "SecondGame")

#novice data
nov.data = subset(data, Condition == "FirstGame")

#expert model
units_exp = exp.data[,c("Condition","UserName")]
conversation_exp = exp.data[,c("Condition","GroupName","ActivityNumber")]
codes_exp = exp.data[,codeCols]
meta_exp = exp.data[,c("CONFIDENCE.Change",
                  "CONFIDENCE.Pre","CONFIDENCE.Post","C.Change")]

set_exp = 
  ena.accumulate.data(
  text_data = exp.data[, 'text'],
  units = units_exp,
  conversation = conversation_exp,
  codes = codes_exp,
  metadata = meta_exp,
  window.size.back = 7,
) |>
  ena.make.set()

set_exp$rotation$rotation.matrix
set_exp$model$points.for.projection

#novice model
units_nov = nov.data[,c("Condition","UserName")]
conversation_nov = nov.data[,c("Condition","GroupName","ActivityNumber")]
codes_nov = nov.data[,codeCols]
meta_nov = nov.data[,c("CONFIDENCE.Change",
                  "CONFIDENCE.Pre","CONFIDENCE.Post","C.Change")]

set_nov = 
  ena.accumulate.data(
  text_data = nov.data[, 'text'],
  units = units_nov,
  conversation = conversation_nov,
  codes = codes_nov,
  metadata = meta_nov,
  window.size.back = 7,
) |>
  ena.make.set(rotation.set = set_exp$rotation)
```

```{r}
# plot expert model (what we projected into) Using plotting wrapper to save time
plot_exp = ena.plotter(set_exp,
                        points = T,
                        mean = T, 
                        network = T,
                        print.plots = F
                       )

# plot test model (points from test model in training model space)
plot_nov = ena.plotter(set_nov,
                        points = T,
                        mean = T, 
                        network = T,
                        print.plots = F)

plot_exp$plot
plot_nov$plot
```

## Projections in ONA

```{r}
data = ona::RS.data

#expert data
exp.data = subset(data, Condition == "SecondGame")

#novice data
nov.data = subset(data, Condition == "FirstGame")

#shared unit cols  
units = c("UserName","Condition","GroupName") 

#shared code cols
codes = c(
          'Data',
          'Technical.Constraints',
          'Performance.Parameters',
          'Client.and.Consultant.Requests',
          'Design.Reasoning',
          'Collaboration')

#shared hoo
hoo = conversation_rules(
  (Condition %in% UNIT$Condition & GroupName %in% UNIT$GroupName))


#expert accum
accum.exp = contexts(exp.data, units_by = units, hoo_rules = hoo) |>
  accumulate_contexts(codes = codes, 
                      decay.function = decay(simple_window, window_size = 7),
                      return.ena.set = FALSE, norm.by = NULL)
#expert model
set.exp = model(accum.exp)

#novice accum
accum.nov = contexts(nov.data, units_by = units, hoo_rules = hoo) |>
  accumulate_contexts(codes = codes, 
                      decay.function = decay(simple_window, window_size = 7),
                      return.ena.set = FALSE, norm.by = NULL)
#novice model
set.nov = model(accum.nov)

# projecting novice data into expert space
set = model(accum.nov, rotation.set = set.exp$rotation)

ona:::plot.ena.ordered.set(set, title = "novice data into expert space") |> 
  units(
    points = set$points, 
    show_mean = TRUE, show_points = TRUE, with_ci = TRUE) |>
  edges(
    weights = set$line.weights) |>
  nodes(
    self_connection_color = "red",
    node_size_multiplier = 0.6) 
```
