---
title: "Factor Analysis in Education Research Using R"
author:
  - name: "Leonie V.D.E. Vogelsmeier"
  - name: "Mohammed Saqr"
  - name: "Sonsoles López-Pernas"
  - name: "Joran Jongerling"
format: 
  html:
    df-print: "paged"
    fig-align: "center"
---


## Factor Analysis in R 


### Preparation

Load libraries
```{r, message = F, warning=F}
library(lavaan) # install.packages("lavaan")
library(psych) # install.packages("psych")
library(semTools) # install.packages("semTools")
library(effectsize) # install.packages("effectsize")
library(rio) # install.packages("rio")
library(tidyr) # install.packages("tidyr")
library(ggplot2) # install.packages("ggplot2")
library(devtools) # install.packages("devtools")
library(sleasy) # install.packages("combinat"); install_github("JoranTiU/sleasy")
```


#### Reading in the data

```{r, warning=F}
dataset <- import("https://github.com/lamethods/data/raw/main/4_teachersBurnout/2.%20Response.xlsx")
var_names <- colnames(dataset)
```

#### Are the data suited for factor analysis?

```{r, eval = F}
describe(dataset)
```


Sample size
```{r}
nrow(dataset)  
```

Bartlett's test 
```{r}
(cortest.bartlett(R = cor(dataset[, var_names]), n = nrow(dataset))$p.value) < 0.05 
```

Kaiser-Meyer-Olkin (KMO) test
```{r}
KMO(dataset)
```
 
Histogram of the responses to each of the items
```{r,  fig.width=6, fig.height=7}
#| label: fig-4
dataset |> pivot_longer(2:ncol(dataset), 
 names_to = "Variable", values_to="Score") |>
   ggplot(aes(x=Score)) + geom_histogram(bins=6) +
     scale_x_continuous(limits=c(0,6),breaks = c(1,2,3,4,5)) +
     facet_wrap("Variable",ncol = 4,scales = "free" ) + theme_minimal()
```
 
Check missing data
```{r}
colSums(is.na(dataset))
```

#### Setting a Holdout Sample Apart

Holdout
```{r}
set.seed(19)
ind <- sample (c (rep("model.building", 438), rep ("holdout", 438)))
tmp <- split (dataset, ind)
model.building <- tmp$model.building
holdout <- tmp$holdout
```

### Step 1: Exploring the Factor Structure

Parallel analysis

```{r,   fig.width=5, fig.height=4}
#| label: fig-5
fa.parallel(x = model.building[,var_names], fa = "fa")
```

Exploratory factor analysis 
```{r}
EFA <- efa (data = model.building[, var_names], nfactors = 4:6, 
            rotation = "geomin", estimator = "MLR", meanstructure = TRUE)
```
 
Extract and sort the BIC values in ascending order.
```{r}
sort(fitMeasures(EFA)["bic",])
```

Obtain the factor loadings for the fivefactors. 
```{r}
EFA$nf5
```

### Step 2: Building the Factor Model and Assessing Fit

Describe the model
```{r}
CFA_model <-'
#  Regressing items on factors
TSC =~ TSC1 + TSC2 + TSC3 + TSC5
TE  =~ TE1  + TE2  + TE3  + TE5
EE  =~ EE1  + EE2  + EE3  + EE4
DE  =~ DE1  + DE2  + DE3
RPA =~ RPA1 + RPA2 + RPA3 + RPA4

# Correlations between factors
TSC ~~ TE
TSC ~~ EE
TSC ~~ DE
TSC ~~ RPA

TE ~~ EE
TE ~~ DE
TE ~~ RPA

EE ~~ DE
EE ~~ RPA

DE ~~ RPA
'
```

Perform CFA
```{r}
CFA <- cfa(model = CFA_model, data = model.building[,var_names], 
           estimator = "MLR", std.lv = TRUE, meanstructure = TRUE)
```


Fit measures
```{r}
globalFit(CFA)
```

See how much the two matrices deviate for each pair of variables
```{r}
localFit(CFA)
```

Look at the loadings of the final model 
```{r}
inspect(object = CFA, what = "std")$lambda
```

### Step 3: Assessing Generalizability

Assessing the generalizability of the CFA model 
```{r}
CFA_holdout <- cfa(model = CFA_model, data = holdout[,var_names], 
                   estimator = "MLR", std.lv = TRUE, meanstructure = TRUE)
```


Fit measures and their interpretation can again be obtained 
```{r}
globalFit(CFA_holdout)
```

Testing local fit again
```{r}
localFit(CFA_holdout)
```


Look at the loadings of the final model when fitted to the holdout sample.
```{r}
inspect(object = CFA_holdout, what = "std")$lambda
```

